---
title: "An R interface to SciDB"
date: "`r Sys.Date()`"
author: B. W. Lewis, Copyright (C) 2016 by Paradigm4, Inc.
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_depth: 4
    mathjax: null
    css: style.css
vignette: >
  %\VignetteIndexEntry{scidb}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

# SciDB

SciDB is a free open-source database that organizes data in n-dimensional
arrays.  SciDB features include ACID transactions, parallel processing,
distributed storage, and efficient sparse array storage. SciDB arrays consist
of a coordinate system whose coordinates are called _dimensions_, and one or
more values called _attributes_ of distinct types at each coordinate.

The `scidb` package provides two ways to interact with SciDB from R:

* Directly running SciDB queries using the native array functional language (AFL),
  optionally returning results as data frames
* Using the `scidb` class and related methods.

The SciDB array classes facilitate programming large-scale SciDB computation in
R using natural R syntax.  This vignette illustrates using SciDB from R by
example. For more detailed information on the functions described in this
vignette, see the manual pages in the package.


## Installing the package

From CRAN (stable, but may lag features on GitHub by several months):
```{r, eval=FALSE}
install.packages("scidb")
```

From the development repository on GitHub (stable branch):
```{r, eval=FALSE}
devtools::install_github("Paradigm4/SciDBR")
```

From the development repository on GitHub (unstable/experimental branch):
```{r, eval=FALSE}
devtools::install_github("Paradigm4/SciDBR", ref="laboratory")
```

"Stable" means that all CRAN checks and package unit tests pass. We try to
make sure that the `scidb` package works with _all_ versions of SciDB but
we only actively test against the current release version of the database.


## Connecting to SciDB

The `scidbconnect` function establishes a connection to a simple HTTP network
service called shim (https://github.com/Paradigm4/shim)
running on a SciDB coordinator instance.  The function may
be safely called multiple times.  Once a connection is established, connection
information is maintained until a different connection is established or the R
session ends.

The network interface optionally supports SSL encryption and SciDB or HTTP
digest user authentication.  The shim service can be configured to support
either open/unencrypted or encrypted/authenticated ports, or both.

*Connect to SciDB on the default shim port*
```{r, eval=FALSE}
library("scidb")
scidbconnect()
```

*Connect to SciDB on an encrypted port 8083 with example SciDB authentication*
```{r, eval=FALSE}
scidbconnect(host="localhost", port=8083, username="scidbuser", password="test")
```
We recommend using only encrypted/authenticated sessions when communicating
with SciDB over public networks.

SciDB user authentication is only supported in SciDB versions 15.7 and
greater.


## Listing and removing SciDB arrays

The `scidblist` function lists SciDB objects (arrays, instances, etc.),
optionally showing detailed schema information for arrays. Returned results may
be filtered using regular expression-style syntax.

The `scidbremove` function removes a SciDB array, or optionally a set of SciDB
arrays. The function accepts a vector of array names, resulting in the removal
of all the specified arrays. Combine this feature with the regular expression
filtering output of `scidblist` to remove sets of arrays matching the filter.
Array names not associated with a specific R session are protected from easy
removal.  Specify `force=TRUE` as indicated in the warning message to remove
those arrays.

## Running AFL queries

Use the `iquery` function to run arbitrary SciDB AFL queries, optionally
returning results to R as data frames.

```{r, eval=FALSE, echo=TRUE}
iquery("build(<v:double>[i=1:2,2,0, j=1:3,1,0], i*j)", return=TRUE)
```
```
  i j v
1 1 1 1
2 2 1 2
3 1 2 2
4 2 2 4
5 1 3 3
6 2 3 6
```


# The scidb data frame-like class

The `scidb` function returns representations of  SciDB arrays as data
frame-like R objects with class `scidb`. The objects display the SciDB
coordinate dimensions and attributes as columns of a data frame. The `scidb`
function accepts arbitrary SciDB AFL expressions or named SciDB arrays.

```{r, eval=FALSE}
x <- scidb("build(<v:double>[i=1:2,2,0, j=1:3,1,0], i*j)")
print(x)
```
```
SciDB expression  build(<v:double>[i=1:2,2,0, j=1:3,1...
SciDB schema  <v:double> [i=1:2,2,0,j=1:3,1,0] 
  variable dimension   type nullable start end chunk
1        i      TRUE  int64    FALSE     1   2     2
2        j      TRUE  int64    FALSE     1   3     1
3        v     FALSE double    FALSE                
```
The variable `x` is a sort of _SciDB array view_; an un-evaluated
SciDB query expression. The value of `x` is evaluated by SciDB lazily only when
needed or explicitly requested. We can force evaluation and return values from
`x` to R in data frame form using the `iquery` function that we used above:
```{r, eval=FALSE}
iquery(x, return=TRUE)
```
```
  i j v
1 1 1 1
2 2 1 2
3 1 2 2
4 2 2 4
5 1 3 3
6 2 3 6
```

Or, we can equivalently use a special empty-indexing operator `[]` on any
SciDB array object to return its values to R:
```{r, eval=FALSE}
x[]
```
```
  i j v
1 1 1 1
2 2 1 2
3 1 2 2
4 2 2 4
5 1 3 3
6 2 3 6
```

Use the `scidbeval` function to evaluate `scidb` values without returning
results to R. The `scidbeval` function stores the evaluation result into new
named SciDB arrays in the database, including optionally temporary arrays.
```{r, eval=FALSE}
y <- scidbeval(x, temp=TRUE)
print(y)
```
```
SciDB expression  R_array1687f71a392851102980131950
SciDB schema  <v:double> [i=1:2,2,0,j=1:3,1,0] 
  variable dimension   type nullable start end chunk
1        i      TRUE  int64    FALSE     1   2     2
2        j      TRUE  int64    FALSE     1   3     1
3        v     FALSE double    FALSE                
```
Note that the SciDB expression associated with the `y` R variable is now a
named SciDB array (automatically named in this case); compare with the
SciDB expression for `x` above.

SciDB array values associated with R variables are by default tied to R's
garbage collector. When the R variable's contents are garbage-collected by R,
the associated SciDB array is removed:
```{r, eval=FALSE}
yname <- y@name
scidbls(yname)
```
```
[1] "R_array1687f71a392851102980131950"
```
```{r, eval=FALSE}
rm(y)
gc()
scidbls(yname)
```
```
character(0)
```
Note! that we use the `@name` slot of the `scidb` object directly here to
interrogate the underlying SciDB expression associated with the R variable.
See the help page for `scidbeval` for more on this topic, including disabling
automatic garbage collection of SciDB arrays.

## Data upload from R to SciDB

Upload local R variables into named SciDB arrays with the `as.scidb` function.
It's not as efficient as native SciDB bulk-uploading options, but works well to
transfer small amounts of data from R into SciDB.

Note that variable names may be changed to reflect naming conventions in SciDB.

Supported R objects for upload include data frames, matrices and vectors, and
numeric sparse matrices from the Matrix package. **Note** that data frames
are uploaded to SciDB using a TSV-delimited text format but matrices and
vectors are uploaded in binary format.

```{r, eval=FALSE}
i <- as.scidb(iris)
print(i)
```
```
Warning message:
In df2scidb(X, name = name, gc = gc, start = start, ...) :
  Attribute names have been changed

SciDB expression  R_array1687f572177bd1102980131950
SciDB schema  <Sepal_Length:double NULL DEFAULT null,Sepal_Width:double NULL DEFAULT null,Petal_Length:double NULL DEFAULT null,Petal_Width:double NULL DEFAULT null,Species:string NULL DEFAULT null> [row=1:150,150,0] 
      variable dimension   type nullable start end chunk
1          row      TRUE  int64    FALSE     1 150   150
2 Sepal_Length     FALSE double     TRUE                
3  Sepal_Width     FALSE double     TRUE                
4 Petal_Length     FALSE double     TRUE                
5  Petal_Width     FALSE double     TRUE                
6      Species     FALSE string     TRUE                
```
```{r, eval=FALSE}
dim(i)
```
```
[1] 150   6
```

Type conversion is applied as required. Some SciDB types (various non-32 bit
integer sizes) are not available in R.  Some R types (complex, factors, etc.)
are not available in SciDB. Missing values in R are translated to SciDB `NULL`
values and vice-versa. (SciDB has a notion of multiple `NULL` codes, but all
codes are translated to R missing values.)


# Important operations on `scidb` values

This section covers a few of the most important operations on `scidb` array
values in R. With few exceptions, operations on SciDB array objects produce new
SciDB array objects. Since objects are effectively SciDB array views,
compositions of R operations build up increasingly complex SciDB queries.

## Project and slice

Use `project`, or equivalently the data frame single column subset notation
`$` to  project the array object onto a subset of its SciDB attributes.
The general bracket `[` indexing function is not supported, use `project`.

```{r, eval=FALSE}
i <- as.scidb(iris)

project(i, "Species")
```
```
SciDB expression  project(R_array1687f572177bd1102980...
SciDB schema  <Species:string NULL DEFAULT null> [row=1:150,150,0] 
  variable dimension   type nullable start end chunk
1      row      TRUE  int64    FALSE     1 150   150
2  Species     FALSE string     TRUE                
```
```{r, eval=FALSE}
i$Species
```
```
SciDB expression  project(R_array1687f572177bd1102980...
SciDB schema  <Species:string NULL DEFAULT null> [row=1:150,150,0] 
  variable dimension   type nullable start end chunk
1      row      TRUE  int64    FALSE     1 150   150
2  Species     FALSE string     TRUE                
```
```{r, eval=FALSE}
i[, "Species"]
```
```
SciDB expression  project(R_array1687f572177bd1102980...
SciDB schema  <Species:string NULL DEFAULT null> [row=1:150,150,0] 
  variable dimension   type nullable start end chunk
1      row      TRUE  int64    FALSE     1 150   150
2  Species     FALSE string     TRUE                
```

Use the `slice` function to slice out one dimension of a SciDB array
at a specific coordinate value.
```{r, eval=FALSE}
(x <- scidb("build(<v:double>[i=1:2,2,0, j=1:3,1,0], i*j)"))
```
```
SciDB expression  build(<v:double>[i=1:2,2,0, j=1:3,1...
SciDB schema  <v:double> [i=1:2,2,0,j=1:3,1,0] 
  variable dimension   type nullable start end chunk
1        i      TRUE  int64    FALSE     1   2     2
2        j      TRUE  int64    FALSE     1   3     1
3        v     FALSE double    FALSE                
```
```{r, eval=FALSE}
slice(x, "j", 2)
```
```
SciDB expression  slice(build(<v:double>[i=1:2,2,0, j...
SciDB schema  <v:double> [i=1:2,2,0] 
  variable dimension   type nullable start end chunk
1        i      TRUE  int64    FALSE     1   2     2
2        v     FALSE double    FALSE                
```

## Subset

Filter SciDB array values with R's `subset` function in one of two
forms:

* Explicit AFL filter expressions
* Implicit queries generated from R expressions

The implicit form can sometimes generate more efficient queries and
allows you to mix simple R scalar values in the SciDB expression. The
following examples illustrate each form.

1. Explicit AFL filter expression (note the quoted AFL expression)
```{r, eval=FALSE}
y <- subset(i, "Species = 'setosa' and row > 40")
y@name
```
```
[1] "filter(R_array1687f572177bd1102980, Species = 'setosa' and row > 40)"
```

2. Implicit query from R expression (note the unquoted R expression and use of R scalar variable)
```{r, eval=FALSE}
rownum <- 40
z <- subset(i, Species == 'setosa' & row > rownum)
z@name
```
```
[1] "filter(between(R_array1687f572177bd1102980, 41, null), Species = 'setosa' )"
```
See `?subset.scidb` for more details.


## Transform

R's `transform` function acts as a wrapper for the SciDB "apply" operator. With it,
you can add new attributes (variables) to SciDB arrays or replace existing ones with
new values.

The syntax of `transform` is
```
transform(array, new_value1="quoted AFL expression 1", new_value2="quoted AFL expr 2", ...)
```
Let's try it:

```{r, eval=FALSE}
i <- as.scidb(iris)
x <- transform(i, Species_index="iif(Species='virginica', 1, iif(Species='versicolor', 2, 3))")
head(x)
```
```
  row Sepal_Length Sepal_Width Petal_Length Petal_Width Species Species_index
1   1          5.1         3.5          1.4         0.2  setosa             3
2   2          4.9         3.0          1.4         0.2  setosa             3
3   3          4.7         3.2          1.3         0.2  setosa             3
4   4          4.6         3.1          1.5         0.2  setosa             3
5   5          5.0         3.6          1.4         0.2  setosa             3
6   6          5.4         3.9          1.7         0.4  setosa             3
```
See `?transform.scidb` for many more details.

## Redimension

Aggregations and other operations are most efficient along SciDB coordinate axes.
The SciDB "redimension" operator is used to set those up from among existing
dimension and attribute values in a SciDB array or array expression.

Let's illustrate by making the "Species\_index" attribute in the last example a
dimension, repeating the steps in the last example for completeness.
```{r, eval=FALSE}
i <- as.scidb(iris)
x <- transform(i, Species_index="iif(Species='virginica', 1, iif(Species='versicolor', 2, 3))")
y <- redimension(x, dim=c("row", "Species_index"))
print(y)
```
```
SciDB expression  redimension(apply(R_array1ebfe51174...
SciDB schema  <Sepal_Length:double NULL DEFAULT null,Sepal_Width:double NULL DEFAULT null,Petal_Length:double NULL DEFAULT null,Petal_Width:double NULL DEFAULT null,Species:string NULL DEFAULT null> [row=1:150,150,0,Species_index=0:*,6667,0] 
       variable dimension   type nullable start end chunk
1           row      TRUE  int64    FALSE     1 150   150
2 Species_index      TRUE  int64    FALSE     0   *  6667
3  Sepal_Length     FALSE double     TRUE                
4   Sepal_Width     FALSE double     TRUE                
5  Petal_Length     FALSE double     TRUE                
6   Petal_Width     FALSE double     TRUE                
7       Species     FALSE string     TRUE
```

## Aggregate

SciDB can very rapidly aggregate array data along the array coordinate system.
R's standard `aggregate` function is overloaded for `scidb` array objects with
some SciDB-specific extensions.

Let's extend the last few examples to compute the average petal length grouped
by Species\_index. We include the original "Species" variable to help identify
the index. We use the special empty indexing operator to return the result to
R.
```{r, eval=FALSE}
i <- as.scidb(iris)
x <- transform(i, Species_index="iif(Species='virginica', 1, iif(Species='versicolor', 2, 3))")
y <- redimension(x, dim=c("row", "Species_index"))
aggregate(y, by="Species_index", FUN="avg(Petal_Length), max(Species) as Species")[]
```
```
  Species_index Petal_Length_avg    Species
1             1            5.552  virginica
2             2            4.260 versicolor
3             3            1.462     setosa
```
The above is equivalent to the naive R aggregation:
```{r, eval=TRUE}
aggregate(iris$Petal.Length, by=list(iris$Species), FUN=mean)
```

SciDB's version of `aggregate` includes many extras like windowed aggregates
and grouping by SciDB attributes (and future versions will include
even better high-performance ad-hoc grouped aggregates).

SciDB cumulative aggregation is supported by the R `cumulate` function.  See
`?regrid` and `?xgrid` for help with high-performance SciDB coordinate-based
aggregation and interpolation (to perform fast spatial or temporal decimation,
for instance).

See `?'aggregate,scidb-method'` for help and more examples.

## Merge and join

R's `merge` function implements SciDB `join`, `cross_join` and `merge` operators,
as well as some extensions including limited outer joins and joining on SciDB
array attributes.

The following example illustrates the difference between join and merge.
```{r, eval=FALSE}
i <- as.scidb(iris)
a <- i$Petal_Length
b <- i$Petal_Width
merge(a, b)                # SciDB (inner) join
```
```
SciDB expression  join(project(R_array1ebfe5145bfb511...
SciDB schema  <Petal_Length:double NULL DEFAULT null,Petal_Width:double NULL DEFAULT null> [row=1:150,150,0] 
      variable dimension   type nullable start end chunk
1          row      TRUE  int64    FALSE     1 150   150
2 Petal_Length     FALSE double     TRUE                
3  Petal_Width     FALSE double     TRUE                
```
```{r, eval=FALSE}
merge(a, b, merge=TRUE)   # SciDB "merge"
```
```
SciDB expression  merge(project(R_array1ebfe5145bfb51...
SciDB schema  <Petal_Length:double NULL DEFAULT null> [row=1:150,150,0] 
      variable dimension   type nullable start end chunk
1          row      TRUE  int64    FALSE     1 150   150
2 Petal_Length     FALSE double     TRUE                
```

Here is an example of a SciDB "cross join" along one common axis between two
arrays with different dimensions.
```{r, eval=FALSE}
i <- as.scidb(iris)
a <- i$Petal_Length

x <- transform(i, Species_index="iif(Species='virginica', 1, iif(Species='versicolor', 2, 3))")
y <- redimension(x, dim=c("row", "Species_index"))$Petal_Width

print(a)
```
```
SciDB expression  project(R_array1ebfe561ae9cd1102983...
SciDB schema  <Petal_Length:double NULL DEFAULT null> [row=1:150,150,0] 
      variable dimension   type nullable start end chunk
1          row      TRUE  int64    FALSE     1 150   150
2 Petal_Length     FALSE double     TRUE                
```
```{r, eval=FALSE}
print(y)
```
```
SciDB expression  project(redimension(apply(R_array1e...
SciDB schema  <Petal_Width:double NULL DEFAULT null> [row=1:150,150,0,Species_index=0:*,6667,0] 
       variable dimension   type nullable start end chunk
1           row      TRUE  int64    FALSE     1 150   150
2 Species_index      TRUE  int64    FALSE     0   *  6667
3   Petal_Width     FALSE double     TRUE                
```
```{r, eval=FALSE}
merge(a, y)
```
```
SciDB expression  cross_join(project(redimension(appl...
SciDB schema  <Petal_Width:double NULL DEFAULT null,Petal_Length:double NULL DEFAULT null> [row=1:150,150,0,Species_index=0:*,6667,0] 
       variable dimension   type nullable start end chunk
1           row      TRUE  int64    FALSE     1 150   150
2 Species_index      TRUE  int64    FALSE     0   *  6667
3   Petal_Width     FALSE double     TRUE                
4  Petal_Length     FALSE double     TRUE                
```
Many additional options are available, see `?merge.scidb` for examples.

### Array concatenation
Use the `rbind` function to concatenate two SciDB arrays with identical
dimensions and identical numbers of attributes of the same types
along their last dimension. Here  is an example:
```{r, eval=FALSE}
i <- as.scidb(iris)
x <- project(i, c("Petal_Length", "Petal_Width"))
count(x)
```
```
[1] 150
```
```{r, eval=FALSE}
count(rbind(x, x))
```
```
[1] 300
```
Note that the result has an unbounded last SciDB coordinate dimension.

Use the `cbind` function to concatenate attributes across two SciDB arrays with
identical dimensions (coordinate schema).
```{r, eval=FALSE}
x
```
```
SciDB expression  project(R_arraya5d8655c582110299266...
SciDB schema  <Petal_Length:double NULL DEFAULT null,Petal_Width:double NULL DEFAULT null> [row=1:150,150,0] 
      variable dimension   type nullable start end chunk
1          row      TRUE  int64    FALSE     1 150   150
2 Petal_Length     FALSE double     TRUE                
3  Petal_Width     FALSE double     TRUE                
```
```{r, eval=FALSE}
cbind(x,x)
```
```
SciDB expression  join(project(R_arraya5d8655c5821102...
SciDB schema  <Petal_Length:double NULL DEFAULT null,Petal_Width:double NULL DEFAULT null,Petal_Length_1:double NULL DEFAULT null,Petal_Width_1:double NULL DEFAULT null> [row=1:150,150,0] 
        variable dimension   type nullable start end chunk
1            row      TRUE  int64    FALSE     1 150   150
2   Petal_Length     FALSE double     TRUE                
3    Petal_Width     FALSE double     TRUE                
4 Petal_Length_1     FALSE double     TRUE                
5  Petal_Width_1     FALSE double     TRUE                
```
Note that attribute names are adjusted as required to remain unique.

### Antijon
Use the `antijoin` function to return a logical-valued mask of array coordinates
that _do not_ join between two arrays with the same coordinate schema.


# Other SciDB operations

The `scidb` package defines many additional SciDB-related functions besides the
most important few outlined above.


## SciDB schema-related functions

The `cast` function works like the SciDB "cast" operator; use it to rename any
part of a SciDB schema, or use the slightly more convenient `dimension_rename`
and `attribute_rename` functions.

The following functions return scalar or vector values to R (not SciDB array
objects).  The `dimensions` and `scidb_attribute` functions return the names of
the SciDB array dimension and attribute variables, respectively.  The
`scidb_nullable`, `scidb_types`, `scidb_coordinate_bounds`, and `schema`
functions return information about SciDB attributes and array schema.

## SciDB schema/shape-altering operations

We've already seen the very important `redimension` function. Use the
`repart` function similarly to the SciDB "repart" operator to alter the
SciDB array coordinate partitions. Use the `reshape` function to change
the dimensionality of a SciDB array, and similarly the `unpack` function
to flatten a SciDB array into a table.

The `subarray` function is a wrapper for the SciDB "subarray" operator that
filters an array along its coordinates and resets the top-left coordinate to
the origin.

## Filling in missing values

Use the `replaceNA` function to replace missing values in SciDB arrays with a
constant, similarly to the SciDB "substitute" operator (the name "substitute"
already had a substantially different meaning in R, so we picked a different
function name).


## SciDB catalog operations

Use `scidbremove` or, equivalently, `scidbrm` to remove SciDB arrays from the
database (permanently!). Related to this topic, use `scidbeval` to control the
connection of R's garbage collector to SciDB arrays associated with R
variables. The `persist` function can be used to recursively modify the
garbage-collection association for any R `scidb` array object, including all
its dependencies.

The `scidbls` function can be used to list arrays and many other SciDB database
objects.

Use the `remove_old_versions` function to remove all the old versions in the
SciDB catalog of a named array, leaving only the last version.

Use `rename` to rename a stored SciDB array in the database.

The `count` function counts the number of elements in the specified SciDB array
or expression just like the SciDB "op\_count" operator/macro.


# Package options

The following package options can be set with R's `options` function.
The most interesting ones are the last two! Set
`options(scidb.debug=TRUE)` to see interesting debugging information
including all the issued SciDB queries. Set `options(scidb.stream=TRUE)`
to avoid creating server-side data files which is most useful when you're
downloading lots of data from SciDB into R.

```
# The scidb.version option is set during scidbconnect(). However, users
# may carefully override it to enable certain bug fixes specific to older
# versions of SciDB.
  options(scidb.version=15.7)

# Default shim port and host.
  options(scidb.default_shim_port=8080L)
  options(scidb.default_shim_host="localhost")

# Make it harder to remove arrays. When this option is TRUE, users
# have to specify scidbrm(array, force=TRUE) to remove arrays that do not
# begin with "R_array".
  options(scidb.safe_remove=TRUE)

# Disable SSL certificate host name checking by default. This is important mostly
# for Amazon EC2 where hostnames rarely match their DNS names. If you enable this
# then the shim SSL certificate CN entry *must* match the server host name for the
# encrypted session to work. Set this TRUE for stronger security (help avoid MTM)
# in SSL connections.
  options(scidb.verifyhost=FALSE)

# Set to TRUE to enable experimental shim stream protocol, avoids copying query
# output to data file on server # (see https://github.com/Paradigm4/shim).
# This must be set to FALSE on versions of SciDB prior to 15.7.
  options(scidb.stream=TRUE)

# Set to TRUE to enable debugging.
  options(scidb.debug=FALSE)
```

# Special note on the `curl` package dependency

Newer versions of the `scidb` package depend on  Jeroen Ooms' `curl` package,
replacing the older `RCurl` dependency and fixing a number of memory,
performance and stability issues.

As of this writing, the `curl` package has one significant problem. Users are
not able to cancel HTTP requests until data transfers begin. In the context of
SciDB interaction, that means that users can't cancel a SciDB query until data
starts flowing back to R.

There exists an experimental version of the `curl` package that does add
this capability, but at the cost of some additional CPU overhead. If you
want to try that out, you can install it directly from GitHub using the
`devtools` package with:

```{r, eval=FALSE}
devtools::install_github("jeroenooms/curl", ref=devtools::github_pull(55))
```
